Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-02-22 paper.tex
tweaks

% Journal Definitions
\def   \aj {{\rm {AJ}}}
\def   \araa {{\rm {ARA\&A}}}
\def   \apj {{\rm {ApJ}}}
\def   \icarus {{\rm {Icarus}}}
\def   \apjs {{\rm {ApJS}}}
\def   \apss {{\rm {Ap\&SS}}}
\def   \aap {{\rm {A\&A}}}
\def   \aapr {{\rm {A\&AR}}}
\def   \aaps {{\rm {A\&AS}}}
\def   \baas {{\rm {BAAS}}}
\def   \memras {{\rm {MmRAS}}}
\def   \mnras {{\rm {MNRAS}}}
\def   \prl {{\rm {Phys. Rev. Lett.}}}
\def   \jqsrt {{\rm {Journal of Quantitative Spectroscopy and Radiative Transfer}}}
\def   \apjl{\rm {ApJL}}
\def   \nat{\rm {Nat.}}

@@

\begin{keywords}
circumstellar matter -- infrared: stars.
\end{keywords}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012; Joos et al. 2013). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \newline

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources (Andr\'e et al. 2000), finding masses between 0.4 and  $>$1 M$_{\odot}$ (J\o rgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using the Jansky Very Large Array (JVLA) to observe NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). Zapata et al. (2013) used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \newline

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation (Cossins et al. 2010; Ruge et al. 2013). Molecular line emission from massive discs has been simulated by Krumholz et al. (2007), assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., Visser et al. (2009, 2011). As shown by Ilee et al. (2011; hereafter I2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of non-LTE 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

To describe the full physical enviroment which we expect these young discs to occur in, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (Keto \& Caselli 2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere (Bonnor 1956; Ebert 1957) with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Caselli et al. 2012; Keto et al. in preparation). Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \newline

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of Boley (2007) and Boley (2009). The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from Ossenkopf and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}, the abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM, Wilson \& Rood 1994). As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. (2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used is LIME (LIne Modeling Engine; Brinch \& Hogerheijde 2010), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in appendix  \ref{sec:gridding}). At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 (Caselli et al. 2012), scaled so the maximum is 1$\times$10$^{-8}$ (because photo-disassoc and freeze out are in similar places, need ref). The H$_2$CO has a step profile abundance with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{9}$ inside it (Young et al. 2004), OCS has a constant abundance of 1.9$\times$10$^{-9}$ (Ren et al. 2011) and C$^{17}$O uses the CO profile of L1544 of KC2010 divided by 1792. We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

@@

Figure \ref{spectra} shows spectra from the lines simulated, one from the centre of the image and one on each side of the disc along the Y=0 axis. As with all the CASA simulated images, these spectra have a total bandwidth of 938$\,$MHz, giving velocity resolutions of between 418 and 527 ms$^{-1}$. The most notable thing about these spectral features is that they are not symmetric about the disk centre. For example, C17O(3$\rightarrow$2) has a significantly stronger blueshifted absorption, due to the presence of a spiral arm along the same line of sight. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be attributed to the non-axisymmetric nature of the model.

@@

\newpage

>>>>

% Journal Definitions
\def   \aj {{\rm {AJ}}}
\def   \araa {{\rm {ARA\&A}}}
\def   \apj {{\rm {ApJ}}}
\def   \icarus {{\rm {Icarus}}}
\def   \apjs {{\rm {ApJS}}}
\def   \apss {{\rm {Ap\&SS}}}
\def   \aap {{\rm {A\&A}}}
\def   \aapr {{\rm {A\&AR}}}
\def   \aaps {{\rm {A\&AS}}}
\def   \baas {{\rm {BAAS}}}
\def   \memras {{\rm {MmRAS}}}
\def   \mnras {{\rm {MNRAS}}}
\def   \prl {{\rm {Phys. Rev. Lett.}}}
\def   \jqsrt {{\rm {Journal of Quantitative Spectroscopy and Radiative Transfer}}}
\def   \apjl{\rm {ApJL}}
\def   \nat{\rm {Nat.}}
\def   \pasj{\rm {PASJ}}

@@

\begin{keywords}
Stars: Circumstellar Matter -- infrared; Stars: Formation -- infrared; Planetary Systems: Protoplanetary Discs.
\end{keywords}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores \citep[e.g.]{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}. The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. \citealt{Joos2012}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \newline

After the pioneer work of, e.g., \citet{Chandler1995,Brown2000,Looney2000}, recent interferometric observations have discovered compact embedded discs in a sample of Class 0 sources \citep{Andre1999}), finding masses between 0.4 and  $>$1 M$_{\odot}$ \citep{Jorgensen2007,Jorgensen2009,Enoch2011}. A 130\,AU disc was discovered toward a Class 0 source in Perseus, using the Jansky Very Large Array (JVLA) to observe NH$_3$ \citep{Choi2007}. \citet{Pineda2012} observed methyl formate (HCOOCH$_3$) with the Atacama Large Millimetre/sub-millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. \citet{Persson2012} observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets \citep{Hartogh2012}. \citet{Zapata2013} used ALMA to observe a disc of size $\sim$50$\,$AU in the other Class 0 source in IRAS 16293-2422 and detected infall towards this disc as inverse P-Cygni profiles in HCN and CH$_3$OH. When ALMA is completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on the theoretical models mentioned previously. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs. \newline

Simulated observations of gravitationally unstable discs have already been performed to study the continuum emission, measure the structure and investigate possible fragmentation \citep{Cossins2010,Ruge2013}. Molecular line emission from massive discs has been simulated by \citet{Krumholz2007}, assuming local thermodynamic equilibrium (LTE). In this paper, we focus on the self-gravitating discs of \citet{Boley2008}, where episodic heating induced by spiral shocks is present. This may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g., \citet{Visser2009,Visser2011}. As shown by \citet{Ilee2011}, hereafter I2011, the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of non-LTE 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by I2011 and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect.\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sect.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

To describe the full physical enviroment which we expect these young discs to occur in, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 \citep{Keto2010}, hereafter KC2010. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \newline

@@

The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the model of the young protoplanetary disc is used. The disc structure is derived by the hydrodynamic model of \citet{Boley2007,Boley2009}. The particular model considered here (as well as in I2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating Keplerian disc (figure \ref{velocity}) featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The underlying model is sampled as a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$AU in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout the disc and the envelope. The dust opacities were adopted from Ossenkopf and Henning (1994) and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.\newline

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. The abundances of 125 species related by 1334 reactions were calculated through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$AU.  From I2011 we selected the four species which appear to trace different regions of the disc: the inner 20\,AU (OCS), the inner 40\,AU (H$_2$CO),  the region between $\simeq$40 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}, the abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM, \citet{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \cite{Caselli2012}). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the LAMDA (Leiden Atomic and Molecular DAtabase) database (Sch\"oier et al. (2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used is LIME (LIne Modeling Engine; \cite{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in appendix  \ref{sec:gridding}). At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \cite{Brinch2010} to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (because photo-disassoc and freeze out are in similar places, {\bf need ref for this}). The H$_2$CO has a step profile abundance with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{9}$ inside it \citep{Young2004}, OCS has a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011} and C$^{17}$O uses the CO profile of L1544 of KC2010 divided by 1792. We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

@@

Figure \ref{spectra} shows spectra from the lines simulated, one from the centre of the image and one on each side of the disc along the Y=0 axis. As with all the CASA simulated images, these spectra have a total bandwidth of 938$\,$MHz, giving velocity resolutions of between 418 and 527 ms$^{-1}$. The most notable thing about these spectral features is that they are not symmetric about the disk centre. For example, C$^{17}$O(3$\rightarrow$2) has a significantly stronger blueshifted absorption, due to the presence of a spiral arm along the same line of sight. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be attributed to the non-axisymmetric nature of the model.

@@

\bibliographystyle{apalike} 
\bibliography{new_bib}
